---
title: "Analysis of injuries and economical damage from weather events"
output: 
  html_document:
    keep_md: true
---
#Synopsys
In this analysis we calculate which events are most harmfull to human health and which have the biggest economic impact. We start from the raw storm events data from NOAA (available here: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). We do some data cleaning (removing non continental events, cleaning the labels and typos). Afterwards we aggregate the injuries per event type. We also calculate the economic impact per event and aggregate it per event type.
At the end, we present a couple of plots with the top 20 of events with most impact on human health and economy. For more details, the complete per event aggregated data is at the end of the analysis.

#Data Processing
##Load Data
First load the raw data from the bz2 file
```{r cache=TRUE} 
storm_raw<-read.csv("repdata-data-StormData.csv.bz2")
```
##Clean data
In the data cleaning we will apply the following processing in order to filter and consolidate event types. The purpose is to have the same event type even if small differences where introduced during data input.

* Convert all Event Types to uppercase to ease the processing
* Remove Marine and Waterspout events as these do not happen accross US
* Remove leading spaces, ending spaces and ending 'S' characters
* Replace repeating spaces by a single space
* Replace backslash with espace
* Correct Thunderstorm typos

```{r}
storm_clean<-storm_raw
storm_clean$EVTYPE<-toupper(storm_clean$EVTYPE) #Convert all EVTYPE to upper case
storm_clean<-subset(storm_clean,!grepl("^MARINE|WATERSPOUT",storm_clean$EVTYPE)) #Remove Marine type events as analysis is only for accross US
storm_clean$EVTYPE<-gsub("S$| $|^ ","",storm_clean$EVTYPE) #Remove ending S, leading and ending spaces
storm_clean$EVTYPE<-gsub(" +"," ",storm_clean$EVTYPE) #Remove repeating spaces
storm_clean$EVTYPE<-gsub("/"," ",storm_clean$EVTYPE) #Remove backlash (some items have, other have spaces)
storm_clean$EVTYPE<-gsub("T.*STORM|T.*STORMS|TSTM","THUNDERSTORM",storm_clean$EVTYPE) #Correct many types with Thunderstorm
proper=function(x) paste0(toupper(substr(x, 1, 1)), tolower(substring(x, 2)))
storm_clean$EVTYPE<-proper(storm_clean$EVTYPE)
```

##Analysis
###Injuries per event type
We want to have a view on the events with most impact on human health. We will use the "Injuries" column in the data set to measure this impact. We aggregate the results per Event Type. We keep only those for which the number of injuries is greater than zero.
```{r}
#Select only Injuries and Fatalities columns and aggregate per event type
fatalities<-aggregate(
  x=storm_clean[c("INJURIES","FATALITIES")],
  list(storm_clean$EVTYPE),
  sum)
names(fatalities)[1]<-c("EVTYPE")
#Keep only fatalities with number of injuries greater than zero
fatalities<-subset(fatalities,INJURIES>0)
#Order descending
fatalities<-fatalities[with(fatalities,order(-INJURIES,EVTYPE)),]
```

###Total economical damage per event type
The economic impact is derived from columns:

* PROPDMG: Property Damage
* CROPDMG: Crop Damage
* PROPDMGEXP: Property damage exponent
* CROPDMGEXP: Crop damage exponent

We create a multipliers data frame to convert from an exponent (columns PROPDMGEXP and CROPDMGEXP) to a numerical value. For instance, we will convert values such as h, k or m to numbers 100, 1,000 or 1,000,000. This requires some data cleansing on the EXP columns (all to upper case, replace symbols with the same meaning to a single one)
We follow the conversion findings shown in the following article:
 https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html
These exponent multipliers are then used to calculate a Total damage (in $) from the PROPDMG and CROPDMG columns.
```{r}
multipliers<-data.frame(c("H","K","M","B","+","-","?","1"),c(100,1000,1000000,1000000000,1,0,0,10))
names(multipliers)<-c("code","PROPDMGMULT") #Create data frame xwith multipliers
#Keep only columns relevant for this analysis
damages<-storm_clean[c("EVTYPE","PROPDMGEXP","CROPDMGEXP","PROPDMG","CROPDMG")]
damages$PROPDMGEXP<-toupper(damages$PROPDMGEXP)
damages$CROPDMGEXP<-toupper(damages$CROPDMGEXP)
damages$PROPDMGEXP<-sub("[0-9]|^$","1",damages$PROPDMGEXP) #Replace numerics and empty by 1
damages$CROPDMGEXP<-sub("[0-9]|^$","1",damages$CROPDMGEXP) #Replace numerics and empty by 1
 #Lookup multipliers for PROPDMGEXP
damages_merge<-merge(damages,multipliers,by.x=c("PROPDMGEXP"),by.y=c("code"),all.x=TRUE)
names(multipliers)<-c("code","CROPDMGMULT")
#Lookup multipliers for CROPDMGEXP
damages_merge<-merge(damages_merge,multipliers,by.x=c("CROPDMGEXP"),by.y=c("code"),all.x=TRUE)
#Calculate total damage
damages_merge$TOTALDMG=
	damages_merge$CROPDMG*damages_merge$CROPDMGMULT+
	damages_merge$PROPDMG*damages_merge$PROPDMGMULT
 the_na<-subset(damages_merge,is.na(TOTALDMG))
damages_agg<-aggregate(x=damages_merge[c("TOTALDMG")],list(damages_merge$EVTYPE),sum)
#Aggregate per event type
names(damages_agg)[1]<-c("EVTYPE")
#Remove results where damage is zero
damages_agg<-subset(damages_agg,TOTALDMG>0)
#Sort descending
damages_agg<-damages_agg[with(damages_agg,order(-TOTALDMG)),]
```

#Results
##Summary
### Injuries per event type
Following graph shows total number of injuries for top 20 most impacting events. The table in the *Detailed result* section shows all data
```{r}
fatalities_plot<-fatalities[1:20,] #Keep only top 20
#Create ordered factor for labels
fatalities_plot$EVTYPE<-factor(
  fatalities_plot$EVTYPE,
  levels=unique(fatalities_plot$EVTYPE))
linch <-  max(strwidth(fatalities_plot$EVTYPE, "inch")+0.4, na.rm = TRUE)
par(mai=c(1.02,linch,0.82,0.42))
fatalities_plot$INJURIES<-fatalities_plot$INJURIES #Convert to thousands
axis_ticks=10^( #Create logarithmic axis ticks
  seq(
    floor(log10(min(fatalities_plot$INJURIES))), 
      log10(max(fatalities_plot$INJURIES)),
      length.out = 10))
barplot(
        fatalities_plot$INJURIES,
        names=fatalities_plot$EVTYPE,
        las=2,
        log="x",
        xaxt="n",
        xaxt="n",
        main="Total injuries per event type (Top 20)",
        horiz=TRUE,
        xlim=c(min(axis_ticks),max(axis_ticks))
        )
axis(1, #Add axis
     at=axis_ticks,
     las=2,
     labels=prettyNum(
       axis_ticks,
       big.mark=",",
       decimal.mark = ".",
       digits=0,
       scientific=FALSE)
     )
mtext("Number of injured, log scale", side=1, line=4)
```

We see that Tornado,Thunderstorm wind and Flood Events have the biggest impact in terms of Injuries.

##Total economical damages per Event Type
### Ecnomical damage per event type
Following graph shows the economical impact for top 20 most impacting events. The table in the *Detailed result* section shows all data
```{r fig.width=8}
damages_plot<-damages_agg[1:20,] #Keep Top 20
#Create ordered factor
damages_plot$EVTYPE<-factor(
  damages_plot$EVTYPE,
  levels=unique(damages_plot$EVTYPE))
linch <-  max(strwidth(damages_plot$EVTYPE, "inch")+0.4, na.rm = TRUE)
par(mai=c(1.02,linch,0.82,0.42))
damages_plot$TOTALDMG<-damages_plot$TOTALDMG/1000000 #Convert to millions
axis_ticks=10^(seq(
  floor(log10(min(damages_plot$TOTALDMG))), 
  log10(max(damages_plot$TOTALDMG)), 
  length.out = 10))
barplot(
  damages_plot$TOTALDMG,
  names=damages_plot$EVTYPE,
  horiz=TRUE,
  las=2,
  log="x",
  xaxt="n",
  main="Total damages per event type (Top 20)",
  xlim=c(min(axis_ticks),max(axis_ticks)))
axis(1,
     at=axis_ticks,
     las=2,
     labels=prettyNum(
       axis_ticks,
       big.mark = ",",
       digits=0,
       scientific=FALSE)
     )
mtext("Total damage (million $) log scale", side=1, line=4)
```

We see that Flood, Hurricane Tornado and Typhoon have the biggest economic impact.

##Detailed Results
###Total injuries per events type.
```{r results="asis"}
library(xtable)
#Format number in US Locale
fatalities$INJURIES<-prettyNum(fatalities$INJURIES,big.mark=",",decimal.mark=".")
fatalities$FATALITIES<-prettyNum(fatalities$FATALITIES,big.mark=",",decimal.mark=".")
names(fatalities)<-c("Event","Injured","Fatalities")
xt_fatal<-xtable(fatalities)
print(xt_fatal,type="html",include.rownames=FALSE)
```

###Table showing total damage per event type.
```{r results="asis"}
library(xtable)
damages_agg$TOTALDMG<-prettyNum(
  damages_agg$TOTALDMG,
  big.mark=",",
  decimal.mark=".",
  scientific=FALSE)
names(damages_agg)<-c("Event","Damages ($)")
damages_agg$TOTALDMG<-
xt_damages<-xtable(damages_agg)
print(xt_damages,type="html",include.rownames=FALSE,digits=c(0,0))
```