---
title: "Analysis of injuries and economical damage from weather events"
output: 
  html_document:
    keep_md: true
---
#Data Processing
##Load Data
First load the raw data from the bz2 file
```{r cache=TRUE} 
storm_raw<-read.csv("repdata-data-StormData.csv.bz2")
```
##Clean data
In the data cleaning we will apply the following processing in order to filter and consolidate event types. The purpose is to have the same event type even if small differences where introduced during data input.

* Convert all Event Types to uppercase to ease the processing
* Remove Marine and Waterspout events as these do not happen accross US
* Remove leading spaces, ending spaces and ending 'S' characters
* Replace repeating spaces by a single space
* Replace backslash with espace
* Correct Thunderstorm typos

```{r}
storm_clean<-storm_raw
storm_clean$EVTYPE<-toupper(storm_clean$EVTYPE) #Convert all EVTYPE to upper case
storm_clean<-subset(storm_clean,!grepl("^MARINE|WATERSPOUT",storm_clean$EVTYPE)) #Remove Marine type events as analysis is only for accross US
storm_clean$EVTYPE<-gsub("S$| $|^ ","",storm_clean$EVTYPE) #Remove ending S, leading and ending spaces
storm_clean$EVTYPE<-gsub(" +"," ",storm_clean$EVTYPE) #Remove repeating spaces
storm_clean$EVTYPE<-gsub("/"," ",storm_clean$EVTYPE) #Remove backlash (some items have, other have spaces)
storm_clean$EVTYPE<-gsub("T.*STORM|T.*STORMS|TSTM","THUNDERSTORM",storm_clean$EVTYPE) #Correct many types with Thunderstorm
```

##Analysis
###Injuries per event type
We aggregate the results per Event Type. We keep only those for which the number of injuries is greater than zero.
```{r}
fatalities<-aggregate(x=storm_clean[c("INJURIES","FATALITIES")],list(storm_clean$EVTYPE),sum)
names(fatalities)[1]<-c("EVTYPE")
fatalities<-subset(fatalities,INJURIES>0)
```

###Total economical damage per event type
We create a multipliers data frame to convert from an exponent (h,k,m...) to a numerical value.
We follow the analysis shown here https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html
```{r}
multipliers<-data.frame(c("H","K","M","B","+","-","?","1"),c(100,1000,1000000,1000000000,1,0,0,10))
names(multipliers)<-c("code","PROPDMGMULT")
damages<-storm_clean[c("EVTYPE","PROPDMGEXP","CROPDMGEXP","PROPDMG","CROPDMG")]
damages$PROPDMGEXP<-toupper(damages$PROPDMGEXP)
damages$CROPDMGEXP<-toupper(damages$CROPDMGEXP)
damages$PROPDMGEXP<-sub("[0-9]|^$","1",damages$PROPDMGEXP) #Replace numerics and empty by 1
damages$CROPDMGEXP<-sub("[0-9]|^$","1",damages$CROPDMGEXP) #Replace numerics and empty by 1
damages_merge<-merge(damages,multipliers,by.x=c("PROPDMGEXP"),by.y=c("code"),all.x=TRUE)
names(multipliers)<-c("code","CROPDMGMULT")
damages_merge<-merge(damages_merge,multipliers,by.x=c("CROPDMGEXP"),by.y=c("code"),all.x=TRUE)
damages_merge$TOTALDMG=
	damages_merge$CROPDMG*damages_merge$CROPDMGMULT+
	damages_merge$PROPDMG*damages_merge$PROPDMGMULT
 the_na<-subset(damages_merge,is.na(TOTALDMG))
damages_agg<-aggregate(x=damages_merge[c("TOTALDMG")],list(damages_merge$EVTYPE),sum)
names(damages_agg)[1]<-c("EVTYPE")
damages_agg<-subset(damages_agg,TOTALDMG>0)
damages_agg<-damages_agg[with(damages_agg,order(-TOTALDMG)),]
```

#Results
##Summary
### Injuries per event type
Following graph shows total number of injuries for 19 most impacting events, the other category aggregates all other results. The table in the *Detailed result* section shows all data
```{r}
fatalities<-fatalities[with(fatalities,order(-INJURIES,EVTYPE)),]
fatalities_plot<-fatalities[1:20,]
fatalities_plot$EVTYPE[20]="OTHER"
fatalities_plot$INJURIES[20]=sum(fatalities$INJURIES[20:length(fatalities)])
fatalities_plot$FATALITIES[20]=sum(fatalities$FATALITIES[20:length(fatalities)])
#fatalities_plot$EVTYPE<-as.factor(fatalities_plot$EVTYPE,ordered=FALSE)
fatalities_plot$EVTYPE<-factor(fatalities_plot$EVTYPE,levels=unique(fatalities_plot$EVTYPE))
plot(x=fatalities_plot$EVTYPE,y=fatalities_plot$INJURIES,log="y",type="h")
#barplot(fatalities_plot)
```

##Total economical damages per Event Type
### Ecnomical damage per event type
Following graph shows the economical impact for 19 most impacting events, the other category aggregates all other results. The table in the *Detailed result* section shows all data
```{r}
damages_plot<-damages_agg[1:20,]
damages_plot$EVTYPE[20]="OTHER"
damages_plot$INJURIES[20]=sum(damages_agg$TOTALDMG[20:length(damages_agg)])
damages_plot$EVTYPE<-factor(damages_plot$EVTYPE,levels=unique(damages_plot$EVTYPE))
plot(x=damages_plot$EVTYPE,y=damages_plot$TOTALDMG,log="y",type="h")
```

##Detailed Results
###Total injuries per events type.
```{r results="asis"}
library(xtable)
xt_fatal<-xtable(fatalities)
print(xt_fatal,type="html",include.rownames=FALSE)
```

###Table showing total damage per event type.
```{r results="asis"}
library(xtable)
xt_damages<-xtable(damages_agg)
print(xt_damages,type="html",include.rownames=FALSE)
```